### 2021-05-05 doc/thesis/chapters/dsp.tex
another chapter; from paper

Considering the relationship between frequency and angle, as described above, and after some simple substitutions which will not be described into detail here, the equation for a generic frequency modulated wave is the following
\begin{equation}
  s(t) = A_c\ cos \Big( 2 \pi f_c t + 2 \pi k_f \int m(t) dt \Big)
  \label{equ:fm_func}
\end{equation}
where $A_c$ is the amplitude of the resulting FM signal and $k_f$ is the frequency sensitivity factor.
This sensitivity factor has a direct relationship with the modulation index $\beta$.
\begin{equation}
  \beta = \frac{\Delta f}{f_m} = \frac{k_f A_m}{f_m}
\end{equation}
where $f_m$ is the highest existing frequency, or the bandwidth of the information signal.

@@

%%%%%%%%%%%%%%%%%%%%
\subsection{Frequency Band, Channel allocation/distribution}

>>>>

Considering the relationship between frequency and angle, as described above, and after some simple substitutions which will not be described into detail here, the equation for a generic frequency modulated wave is the following
\begin{equation}
  s(t) = A_c\ cos \Big( 2 \pi f_c t + 2 \pi k_f \int m(t) dt \Big)
  \label{equ:fm_func}
\end{equation}
where $A_c$ is the amplitude of the resulting FM signal and $k_f$ is the frequency sensitivity factor.
This sensitivity factor has a direct relationship with the modulation index $\beta$.
\begin{equation}
  \beta = \frac{\Delta f}{f_m} = \frac{k_f A_m}{f_m}
\end{equation}
where $f_m$ is the highest existing frequency, or the bandwidth of the information signal.
% More info on modulation index: https://fas.org/man/dod-101/navy/docs/es310/FM.htm

@@

%%%%%%%%%%%%%%%%%%%%
\subsection{Frequency Band}

The frequency band that is used for FM broadcasting is defined worldwide and spans from 87.5 to 108 MHz.
However, some countries only partially use the band \cite{ref_itu_regulations}.
The range is located within the so-called Very-High-Frequency (VHF) band, which is open to the public in terms of usage.
This means, that any transmitter may use the specified frequency range freely.
Austria allowed the legal usage in 2006 \cite{ref_austria_rundfunkgesetz_2006}.
However, the transmission power needs to be limited, so that neighboring receivers are not disturbed in receiving any existing channels.
The European Commision specifies the power limit as 50 nW of effective radiated power (ERP) \cite{ref_eu_commision_radio_spectrum}.
Another limitation is, that a transmitter must be capable to transmit on multiple center frequencies within the broadcasting band.
It must not be fixed to a single center frequency.
Through this specification, a transmitter is capable of switching its transmission center frequency, in order to avoid interference with another transmitters signal.

In a practical usecase this means that a transmitter needs to select a channel or frequency range, that is not already occupied by an official transmitter.
If a free range is found, the transmitter may start sending an FM signal with the mentioned power limitations \cite{ref_ebu_fm_modulators_in_europe_power_regulations}.

%https://www.ris.bka.gv.at/GeltendeFassung.wxe?Abfrage=Bundesnormen&Gesetzesnummer=20008807

%%%%%%%%%%%%%%%%%%%%
\subsection{Frequency Spectrum of a Broadcast FM Channel}

The entire FM broadcasting band from 87.5 to 108 MHz divides into multiple channels which can be used.
In Europe, the European Telecommunications Standards Institute (ETSI) sets the respective standards for the usage of these frequencies.
The main specifications for a single FM broadcasting channel are explained in this section.\\

Each channel may allocate a bandwidth of 200 kHz.
Within one of these channels, the maximum deviation from the center frequency shall not exceed 75 kHz.
This leaves a guard band of 25 kHz on either side, to minimize interference with adjacent channels.
Center frequencies may be allocated on multiples of 100 kHz.
However, neighboring channels need to be considered therefor, since this may result in an overlap in the spectrum.\\

Fig.\ref{fig_channel_baseband_freqs} shows the frequency spectrum of a single channel, located at its carrier- or center frequency.
For a better overview, only the positive spectum part is shown.
The negative part is an exact mirror, because the transmitted and received signals from the antenna are real signals - they do not have an imaginary part.
The spectrum consists of multiple parts, which are described in the following paragraphs.

\begin{figure}[!h]
  \centering
    \includegraphics[width=7cm]{img/fm-channel-baseband.png}
  \caption{Allocation of frequencies in an FM channel \cite{ref_fig_channel_freqs}}
  \label{fig_channel_baseband_freqs}
\end{figure}

\subsubsection{Mono Audio Part}

The mono audio stream is located between 30 Hz and 15 kHz.
This signal is built by the sum of the left and right audio channels.
%The lower limit at 30 Hz is to prevent the transmission of a direct current (DC) part, which would require a large amount of power in transmission.
%This effect is one of the downsides of amplitude modulation (AM), where the carrier is located at the channels' center frequeny and requires a high transmission power. % TODO: check this fact
The upper limit at 15 kHz is chosen to maintain a sufficient spacing to the first subcarrier.
For audio streaming, as in FM broadcasting, this is not a limitation, since it is unlikely to have frequencies higher than 15 kHz in an audio stream.
Also, this already reaches the upper limit of the human ears' bandwith, so higher spectral parts will not be heard anyway.

\subsubsection{Pilot Tone}

The first subcarrier is allocated at an offset of 19 kHz from the center frequency.
It is also called the 'pilot tone', since it is a continuous signal.
The pilot tone is used to be able to demodulate the left and right channel of an audio signal correctly, to generate stereo audio.

\subsubsection{Stereo Audio Difference}

A signal that is constructed by the difference between left and right audio channel is modulated on a 38 kHz subcarrier.
This subcarrier is an integer multiple of 19 kHz for practical reasons.
However, the 38 kHz carrier is suppressed and thus not visible in the received spectrum.
It can still be recovered, since it is phase coherent with the 19 kHz pilot tone per definition.
The bandwidth for this difference-signal spans 15 kHz on either side of the subcarrier.
It is used to generate a stereo audio signal, in combination with the mono signal.
This process is explained into more detail in chapter \ref{sec_fm_sig_demod}.

\subsubsection{Additional Services}

Considering all these parts in the spectrum, there is still bandwidth free to use, up to the maximum channel bandwitdh of 100 kHz in one sideband.
Because of that, additional services were added to the pure audio transmission, to provide additional data services and information.
Services that were implemented are the Data Radio Channel (DARC), which is mostly used in Japan and the USA, the Subsidary Communication Authorization (SCA) and the Radio Data System (RDS) \cite{ref_rohde_u_schwarz}.
Out of these, RDS is the most significant service in Europe.
It is used to transmit additional information about the channel, such as the radio stations' name, currently playing songs' title or traffic information.\\

\noindent
The information about the frequency spectrum was taken from \cite{ref_fm_broadcast_tutorial_and_basics}.
